Developing effective theories for the square lattice Hubbard model using density matrix downfolding

### 2019-07-29 draft.tex
cleaned preliminary work a lot based on CuO paper edits I just made. Need to cut 18 -> 13 pages. Planning 3, 6, 4 pages for intro, preliminary work, and proposed work. or 4, 5, 4, dunno

\subsection{Fixed-node diffusion Monte Carlo (FN-DMC)}
Diffusion Monte Carlo (DMC) is a quantum Monte Carlo method which projects out the ground state of a real-space Hamiltonian given some initial trial wave function.
Consider a trial wave function $|\Psi_T\rangle$ and the Hamiltonian $\hat{H}$ with ground state $|\Phi_0\rangle$. 
Applying the projector $e^{-\tau \hat{H}}$ as $\tau \rightarrow \infty$ to $|\Psi_T \rangle$
\begin{equation}
\lim_{\tau \rightarrow \infty} e^{-\tau \hat{H}} |\Psi_T\rangle 
\equiv \lim_{\tau \rightarrow \infty} |\Psi_\text{DMC}(\tau)\rangle \propto \langle \Phi_0|\Psi_T\rangle |\Phi_0\rangle,
\end{equation}
projects out the ground state as long as the trial wave function is not orthogonal to the ground state. 
DMC provides a stochastic implementation of this projector method and involves moving samples from the trial function $\Psi_T(R)$ using the Green function $G(R, R^\prime, \tau) = \langle R | e^{-\tau(\hat{H} - E_T)} | R^\prime \rangle$. 
Since $\hat{H} = \hat{T} + \hat{V}$, kinetic and potential energy terms, the Green function is approximated by a Trotter expansion $G(R, R^\prime, \tau) = \langle R | e^{-\tau(\hat{H} - E_T)} | R^\prime \rangle \sim \Big[e^{-d\tau(\frac{V(R) + V(R^\prime)}{2} - E_T)} \langle R| e^{-d\tau\hat{T}}|R^\prime \rangle + O(d\tau^2) \Big]^N $ where $d\tau = \tau/N$.
This expansion can be interpreted as an interative procedure where samples are moved N times with a small timestep $d\tau$ until convergence.
The constant $E_T$ is a trial energy used to control the normalization of $\Psi_\text{DMC}(\tau, R)$ and is updated at each move.

@@

\section{Preliminary Work}
\subsection{Non-orthogonal determinants in FN-DMC trial wave functions}
In order to become acquainted with our code QWalk \cite{WAGNER20093390} and QMC algorithms in general I worked on implementing and testing multi-Slater-Jastrow trial functions with optimized non-orthogonal determinants (MSJ+NO) in FN-DMC \cite{Pathak2018}.
We assessed the efficiency and compactness of this new trial function by calculating the ground state energy and single particle densities of a C$_2$ molecule using FN-DMC and comparing to the results when using multi-Slater-Jastrow trial functions with optimized orthogonal determinant trial functions (MSJ+O). 
The workflow involved constructing the un-optimized trial wave functions, optimizing the parameters using an energy optimization method \cite{Toulouse2007}, and finally using the optimized trial functions in both variational Monte Carlo (VMC) and FN-DMC calculations. 
We found the average improvement in variational energy when using optimized orthogonal determinants is $\langle E_\text{MSJ} - E_\text{MSJ+O} \rangle = $ 0.32 eV with a standard deviation of 0.08 eV, with an additional improvement from the non-orthogonal determinants of 0.031 eV with a standard deviation of 0.021 eV.
The average improvement in the FN-DMC energy with the MSJ+O trial wave functions was 0.14 eV (standard deviation 0.03 eV) with an additional reduction of 0.032 eV (standard deviation 0.019 eV) when using the MSJ+NO trial wave function.
While the benefit using the MSJ+NO trial wave function seems small, we saw that the FN-DMC energy calculated using an MSJ+NO trial function with only 24 determinants was lower than the FN-DMC energy using an MSJ+O trial function with 55 determinants.
Further, the FN-DMC charge density calculated using MSJ+NO trial functions had stronger bonding character than when using MSJ+O trial functions, a reasonable result as introducing correlations into trial functions allows for electrons to avoid each other while still occupying the same bonding region. 
Our results indicated that using non-orthogonal determinants may lead to more compact multi-Slater-Jastrow trial wave functions for small molecules.

\subsection{Effective theory for CuO molecule using DMD}
As a first step towards developing accurate models for extended systems like solids using DMD, we constructed a many-body effective model for the CuO molecule which accurately describes the eigenstates and spectra seen in experiment up to 2eV above the ground state.
The neutral CuO molecule has been a subject of intense theoretical and experimental study due to the complex structure of its low-energy space.
While the low-energy excitations in the doublet sector of the CuO molecule are well understood, singlet selection rules for spectroscopic measurements and the doublet ground state leave the quartet sector of excitations unexplored.
We construct our $\mathcal{LE}$ based on recent anion photoelectron spectroscopy (APES) measurements of the low-lying excited states of the nuetral CuO molecule. 
The lowest lying excitations involve electrons in the Cu 3d, Cu 4s and O 2p orbitals.
Given that the lowest energy excitations contain a fully-filled Cu 3d shell or a single hole in this shell we define our low-energy space as:
\begin{equation}
\mathcal{LE} = \text{Span(}\{ |\Psi \rangle | \langle \Psi | \hat{n}_{3d} | \Psi \rangle \ge 9,\ \hat{H}_\text{ab}|\Psi\rangle = E |\Psi\rangle \}\text{)}.
\label{eq:LE}
\end{equation}

Our sample space was generated by using multi-Slater-Jastrow trial wavefunctions in FN-DMC.
The parameterization for our low-energy trial wave functions was
\begin{equation}
e^{J}\sum_{i} c_i|\text{D}_i\rangle = \sum_{i} c_i (e^J |\text{D}_i \rangle) \xrightarrow{FN-DMC} \ \sim \sum_i c_i |\Phi_i \rangle \in \mathcal{LE}
\label{eq:sampling}
\end{equation}
where $e^J$ is a three-body Jastrow factor and $|\text{D}_i\rangle$ is a Slater determinant whose nodes accurately approximate those of an eigenstate $|\Phi_i \rangle \in \mathcal{LE}$.
Under the FN-DMC projection these trial wave functions map to states which are very close to linear combinations of eigenstates in $\mathcal{LE}$.
We used symmetry-targeted unrestricted Kohn-Sham (UKS) to generate our determinants $|\text{D}_i\rangle$ using a B3LYP functional, a Trail-Needs pseudopotential, and the VTZ Trail-Needs basis using the package PySCF.
This method allows us to access almost every excitation in $\mathcal{LE}$, except the few cases when two states have identical S$_z$ and number of electrons per irrep.
An example would be the ground state and the state with an excitation from Cu 3d$_{xz} \rightarrow $  O 2p$_x$, where the latter would be inaccessible.
The FN-DMC calculations were conducted with T-moves to make the calculation variational while using pseudopotentials, and a timestep of $\tau = 0.01$.
The coefficients $\{c_j\}$ for our sample states were chosen via a shell-sampling method. 
We begin by fixing the coefficient $c_0 = \sqrt{w}$ where $w \in \{1.0, 0.8,..., 0.2\}$. 
For each choice of $w$ we sample n = 5 states by randomly selecting the unassigned coefficients from a uniform distribution such that $\sum_i c_i^2 = 1$. 
We can then loop over all $i$ to generate shells of decreasing distance near each eigenstate $|\Phi_i\rangle$.

Parameterizing our effective model requires selecting a set of descriptors from which we would like to build our model and a basis on which to build our descriptors.
Occupation energies of and hybridization between Cu 3d, Cu 4s and O 2p are necessary to describe the low-energy excitations in CuO and are best represented by the occupation energies in a molecular orbital basis (MO) which would neatly package both effects.
The lack of doubly occupied Cu 4s states in the low-energy spectrum of CuO and the existence of a large Hund's coupling between the Cu 4s and 3d orbitals in the Cu atom indicate the necessity for a Hubbard U$_s$ and Hund's J$_{sd}$ respectively. 
These two interaction terms are best represented in a localized intrinsic atomic orbital (IAO).
To account for orbital relaxation between excited states in the CuO molecule we will consider including additional hybridization between MOs.
In total our space of canditate models is then 
\begin{equation}
\begin{split}
\{\bar{n}_{4s}, \bar{n}_{p_\pi}, \bar{n}_{p_z}, \hat{n}_{4s\uparrow} \hat{n}_{4s\downarrow},\sum_{i \in \{xy, xz,...\}}\vec{S}_{4s}\cdot \vec{S}_{d_i}\} \ + \\
\text{P}(\bar{c}_{d_\pi}^\dagger \bar{c}_{p_\pi} + h.c.,\ \bar{c}_{d_z^2}^\dagger \bar{c}_{p_z} + h.c.,\ \bar{c}_{4s}^\dagger \bar{c}_{p_z} + h.c.,\ \bar{c}_{d_z^2}^\dagger \bar{c}_{4s} + h.c.)
\end{split}
\label{eq:models}
\end{equation}
where P denotes the power set and operators are defined on the IAO basis unless denoted by a bar, in which case the operator is built on the MO basis. 
The coefficients for each term above will be denoted $\bar{\epsilon}_{4s},\ \bar{\epsilon}_{p_\pi},\ \bar{\epsilon}_{p_z},\ U_s,\ J_{sd},\ \bar{t}_\pi,\ \bar{t}_{dz},\ \bar{t}_{sz},\ \bar{t}_{ds}$. 
All energies will be relative to $\bar{\epsilon_{3d}} = 0$. 
The symbol $\pi$ denotes a contraction over $x, y$ for p orbitals and $xz,\ yz$ for d orbitals. The symbol $\delta$, introduced later, denotes a contraction over $xy,\ x^2-y^2$ for d orbitals.

After fitting each of the potential models in \eqref{eq:models} using ordinary least squares (OLS) linear regression we find three models which describe the energy functional on our sample set accurately with low variance, but whose eigenstates and spectra differ as a consequence of intruder states below an energy of 2eV as illustrated in Figure \ref{fig:InitED}. 
All three models have a first excited state around 1eV, but only in the minimal model does this state fall near the sample set used for the fitting.
The large distance of the eigenstates for the other two models from our sample set indicates that their predicted eigenvalues may suffer from large extrapolation errors, and as such we will call them \textit{intruder} states.
In Figure \ref{fig:Intruder} we present all eigenstates among the three potential models which we classify as intruders using a k-nearest neighbor approach with k = 5.

\begin{figure}[H]
	\centering
	\subfigure[Results of exact diagonalization showing the energies and properties of the first twenty eigenstates for three candidate models fit using ordinary linear regression. The ground state and first excited state are colored and have enlarged markers for clarity. Error bars are 95\% CI calculated by a bootstrap estimate.]{
	 	\label{fig:InitED}
		\includegraphics[width=0.45\textwidth]{figs/init_ed.pdf}
	}
	\qquad
	\subfigure[Collection of unique intruder states from our three potential models selected using a k-nearest neighbor approach with k = 5.]{	
		\label{fig:Intruder}
		\includegraphics[width=0.45\textwidth]{figs/intruder.pdf}
	}
	\caption{Results of OLS fitting on sampled data and selected intruder states.}
\end{figure}

Given that our intruder states correspond to states within $\mathcal{LE}$ but outside our sample set, they must live within the span of the eigenstates which were excluded in our sampling scheme as described before.
This is evidenced by the fact that two of the intruder states correspond explicitly to excitations we could not access: Cu 3d$_{z^2} \rightarrow $ O 2p$_\pi$ and  Cu 3d$_\pi \rightarrow $ O 2p$_\pi$.
We can approximate these states by single rigid MO excitations above the UKS ground state, which results in states with a FN-DMC energy above 4eV.
Including a generous 2eV reduction in FN-DMC energy due to orbital relaxation, we estimate that these two states should lie at least 2eV in energy. 
Since all the other inaccessible eigenstates are double or higher order excitations, we assert a prior: intruder states should lie above 2eV in energy in our effective theories, enforced via an augmented cost function
\begin{equation}
\text{Cost} = \sum_{i} (E_\text{eff}[\Psi_i] - E_\text{ab}[\Psi_i])^2 + \lambda \sum_{p}\text{QHL}(2 - E_\text{eff}[\Psi_p]),\ \text{QHL}(x) = \Theta(x)x^2
\label{eq:cost}
\end{equation}
where $\lambda>0$ is a parameter which can be varied, QHL is a quadratic hinge loss, $\Theta$ is the Heaviside step function, the index $i$ is over our sampled states and $p$ over the selected intruder states.

\begin{figure}[H]
	\centering
	\subfigure[On the left, scores of the three candidate models at various $\lambda$ when fitting our effective theories using the cost function \eqref{eq:cost}. On the right we show predicted versus calculated energies for our low-energy sampled states for the minimal model fit at $\lambda = 20$.]{
	 	\label{fig:Prior}
		\includegraphics[width=0.45\textwidth]{figs/prior_and_regr.pdf}
	}
	\qquad
	\subfigure[Results of exact diagonalization showing the energies and properties of the first thirty eigenstates for three candidate models using linear regression with cost function \ref{eq:cost} and $\lambda = 20$. The ground state and first excited state are colored and have enlarged markers for clarity. Error bars are 95\% CI calculated by a bootstrap estimate.]{	
		\label{fig:FinalED}
		\includegraphics[width=0.45\textwidth]{figs/final_ed.pdf}
	}
	\caption{Results of linear regression using augmented cost function \eqref{eq:cost}.}
\end{figure}

Using our new cost function we find a set of models which accurately describe our sample data, have very similar spectra and eigenvectors to eachother and experiment, and do not contain intruder states below the 2eV prior. 
Shown in Figure \ref{fig:Prior} are the R$^2$ scores and QHL for the three candidate models fit at $\lambda \in [0,20]$. 
We believe that all three models at $\lambda = 20$ should be considered good models as they satisfy our prior and also explain our sample set accurately. 
Also shown are $E_\text{eff}, E_\text{ab}$ for our samples using the minimal model at $\lambda = 20$ to illustrate the high fit quality even with a large $\lambda$.
Figure \ref{fig:FinalED} presents the properties and eigenvalues of the lowest 30 eigenstates for the three fit models at $\lambda = 20$. 
The additional constraint from asserting a prior has seemingly caused a single effective theory to emerge from our set of candidate models.
This conclusion is verified by looking at the values of the fit parameters for our three models, presented in Table 1.
All occupation energies are relative to $\epsilon_{d_\delta}$.
In the doublet sector the ground and first excited state of any of the three models match experiment, as well as the block of states at 2eV corresponding to excitations out of the Cu 3d shell.
The lowest energy quartet states are at 1, 2eV respectively and are in line with the observation of a quartet state at 2eV in APES measurements.

\begin{table}[H]
\begin{center}
\begin{tabular}{l|lllll}
Model &$\epsilon_{d_{z^2}}$ & $\epsilon_{d_\pi}$ & $\epsilon_{4s}$ & $\epsilon_{p_\pi}$ & $\epsilon_{p_z}$ \\ \hline 
Min & 0.32(2)& 0.19(1)& 2.2(4)& 1.7(3)& 0.96(4)\\
Min $ +\ \bar{c}_{d_\pi}^\dagger \bar{c}_{p_\pi}$& 0.32(1)& 0.10(2)& 2.23(3)& 1.8(2)& 0.97(3)\\
Min $ +\ \bar{c}_{d_{z^2}}^\dagger \bar{c}_{p_z}$& 0.29(2)& 0.19(1)& 2.19(5)& 1.7(3)& 0.99(5)\\
\end{tabular} \\

\begin{tabular}{l|llllll}
Model &$t_\pi$ & $t_{dz}$ & $t_{sz}$ & $t_{ds}$ & $J_{sd}$ & $U_s$ \\ \hline 
Min &  -0.57(1)& 0.55(3)& 0.87(2)& 0.44(1)& -0.63(9)& 3.8(2)\\
Min $ +\ \bar{c}_{d_\pi}^\dagger \bar{c}_{p_\pi}$& -0.45(2)& 0.56(2)& 0.89(2)& 0.45(1)& -0.78(7)& 3.7(2)\\
Min $ +\ \bar{c}_{d_{z^2}}^\dagger \bar{c}_{p_z}$& -0.57(1)& 0.54(2)& 0.85(2)& 0.45(2)& -0.58(3)& 3.9(2)\\
\end{tabular} \\

\end{center}
Table 1: Parameters in eV for our three potential models when fit using \eqref{eq:cost} at $\lambda = 20$, energies relative to $\epsilon_{d_\delta} = 0$. Error bars are 95\% CI calculated using a bootstrap estimate.
\end{table}

@@

The first system I will develop a model for is the benzene molecule.
The benzene molecule has chemical formula $C_6 H_6$ and contains a hexagonal ring of carbon atoms with hydrogen atoms sticking out from each carbon.
These carbon atoms are linked by sp$^2$ hybridized $\sigma$ bonds and an alternating sequence of $\pi$ bonds from the C 2p$_z$ orbitals.
Intuitively, the ground state of the system can be seen as a superposition of two configurations of alternating $\pi$ bonds as seen in Figure \textbf{ref} rotated 30$^o$ relative to each other.

@@

The last systems I will develop models for are AA and AB stacked BLG.
AA stacked BLG is simply two sheets of graphene placed on top of one another where each atom is aligned with the atom beneath it.
In AB stacked BLG, a carbon atom from the lower layer is present in the center of each carbon hexagon of the upper layer, and vice versa.
As such, one can think of AB BLG as a shifted version of AA BLG, where the shift is half the size of a single carbonic hexagon.

>>>>

\subsection{Fixed-node diffusion Monte Carlo (FN-DMC)}
Diffusion Monte Carlo (DMC) is a quantum Monte Carlo method which projects out the ground state of a real-space Hamiltonian given some initial trial wave function.
Consider a trial wave function $|\Psi_T\rangle$ and the Hamiltonian $\hat{H}$ with ground state $|\Phi_0\rangle$. 
Applying the projector $e^{-\tau \hat{H}}$ as $\tau \rightarrow \infty$ to $|\Psi_T \rangle$
\begin{equation}
\lim_{\tau \rightarrow \infty} e^{-\tau \hat{H}} |\Psi_T\rangle 
\equiv \lim_{\tau \rightarrow \infty} |\Psi_\text{DMC}(\tau)\rangle \propto \langle \Phi_0|\Psi_T\rangle |\Phi_0\rangle,
\end{equation}
projects out the ground state as long as the trial wave function is not orthogonal to the ground state. 
DMC provides a stochastic implementation of this projector method and involves moving samples from the trial function $\Psi_T(R)$ using the Green function $G(R, R^\prime, \tau) = \langle R | e^{-\tau(\hat{H} - E_T)} | R^\prime \rangle$. 
Since $\hat{H} = \hat{T} + \hat{V}$, kinetic and potential energy terms, the Green function is approximated by a Trotter expansion 
$$G(R, R^\prime, \tau) = \langle R | e^{-\tau(\hat{H} - E_T)} | R^\prime \rangle \sim \Big[e^{-d\tau(\frac{V(R) + V(R^\prime)}{2} - E_T)} \langle R| e^{-d\tau\hat{T}}|R^\prime \rangle + O(d\tau^2) \Big]^N $$ 
where $d\tau = \tau/N$.
This expansion can be interpreted as an interative procedure where samples are moved N times with a small timestep $d\tau$ until convergence.
The constant $E_T$ is a trial energy used to control the normalization of $\Psi_\text{DMC}(\tau, R)$ and is updated at each move.

@@

\section{Preliminary Work}
\subsection{Non-orthogonal determinants in FN-DMC trial wave functions}
In order to become acquainted with our code QWalk \cite{WAGNER20093390} and QMC algorithms in general I worked on implementing and testing multi-Slater-Jastrow trial functions with optimized non-orthogonal determinants (MSJ+NO) in FN-DMC \cite{Pathak2018}.
The MSJ+NO trial wave functions take the form
\begin{equation}
\Psi=e^{J(\vec{\alpha})} \sum_I e^{\hat{\Theta}(\vec{\theta_I})} C_I |D_I (\{ \phi\})\rangle
\end{equation}
where the determinants $|D_I\rangle$ are composed of single particle orbitals from a self-consistent field calculation, $e^J$ is a spin-independent Jastrow factor, $e^{\hat{\Theta}}$ is a spin-dependent orbital rotation matrix, and the wave function has variable parameters $\vec{\alpha}, \vec{\theta}, \vec{C}$ which can be optimized.
We assessed the efficiency and compactness of this new trial function by calculating the ground state energy and single particle densities of a C$_2$ molecule using FN-DMC and comparing to the results when using multi-Slater-Jastrow trial functions with optimized orthogonal determinant trial functions (MSJ+O). 
The workflow involved constructing the un-optimized trial wave functions, optimizing the parameters using an energy optimization method \cite{Toulouse2007}, and finally using the optimized trial functions in both variational Monte Carlo (VMC) and FN-DMC calculations.

We found the average improvement in variational energy when using optimized orthogonal determinants is $\langle E_\text{MSJ} - E_\text{MSJ+O} \rangle = $ 0.32 eV with a standard deviation of 0.08 eV, with an additional improvement from the non-orthogonal determinants of 0.031 eV with a standard deviation of 0.021 eV.
The average improvement in the FN-DMC energy with the MSJ+O trial wave functions was 0.14 eV (standard deviation 0.03 eV) with an additional reduction of 0.032 eV (standard deviation 0.019 eV) when using the MSJ+NO trial wave function.
While the benefit using the MSJ+NO trial wave function seems small, we saw that the FN-DMC energy calculated using an MSJ+NO trial function with only 24 determinants was lower than the FN-DMC energy using an MSJ+O trial function with 55 determinants.
Further, the FN-DMC charge density calculated using MSJ+NO trial functions had stronger bonding character than when using MSJ+O trial functions, a reasonable result as introducing correlations into trial functions allows for electrons to avoid each other while still occupying the same bonding region. 
Our results indicated that using non-orthogonal determinants may lead to more compact multi-Slater-Jastrow trial wave functions for small molecules.

\subsection{Effective theory for CuO molecule using DMD}
As a first step towards developing accurate models for extended systems like solids using DMD, I constructed a many-body effective model for the CuO molecule which accurately describes the lowest-lying eigenstates seen in experiment.
The neutral CuO molecule has been a subject of intense theoretical and experimental study due to the complex structure of its low-energy space.
Unlike lighter transition metal oxide molecules like TiO or FeO, the energy scales related to the active Cu 3d, 4s and O 2p orbitals are very similar.
This leads to many low-energy eigenstates below $\sim $ 3eV which have radically different properties.
As such, no accurate low-energy effective model for the molecule has been constructed successfully.

In the following four sections I will detail the steps I took towards developing a model for the CuO molecule using DMD.
First, I discuss how low-energy states were sampled for the model regression.
This is followed by details of descriptor selection and single particle basis construction.
With the sampled data and candidate descriptors in hand, the feature selection and model fitting procedures are described.
I will conclude by comparing the solutions of the model Hamiltonian to experiment and previous theoretical calculations.

\subsubsection{Sampling low-energy states}
For the CuO molecule, low-energy states were generated by using FN-DMC projected multi-Slater-Jastrow trial wave functions.
The trial wave functions took the form
\begin{equation}
|\Psi_T(\vec{p}) \rangle =  e^{J}\sum_{j} p_j|\text{D}_j\rangle,
\label{eq:sampling}
\end{equation}
where $e^J$ is a Jastrow factor, $|\text{D}_j\rangle$ are selected determinants which we discuss below, and $p_j$ are variable coefficients.
A single three-body Jastrow factor was used for all sampled states optimized on the lowest energy determinant $|\text{D}_0 \rangle$.
The choice to use a single Jastrow factor significantly reduced the computational cost of the DMD method and is justified by noting that large variations in the Jastrow factor would move the state outside the low-energy space.

The determinants $|\text{D}_j \rangle$ were constructed from the orbitals of independent symmetry-targeted unrestricted Kohn Sham (UKS) calculations.
This choice was made to ensure that the three major low-energy variations seen in the molecule can be captured in our samples: occupancy, relaxation and spin configuration of the Cu 3d, 4s and O 2p orbitals.
The symmetry-targeting allows us to select the number of electrons in both the spin-up and spin-down channel for a given irreducible representation of the symmetry group for CuO, letting us generate determinants with variations in both total spin and orbital occupation.
Orbital relaxation effects are captured by having each determinant, corresponding to a unique orbital and spin configuration, be the result of an independent self-consistent field calculation.
The UKS calculations were done using PySCF, a B3LYP functional, Trail-Needs psuedopotential and VTZ Trail-Needs basis.

Of these UKS determinants, only those which contribute to accurately describing the low-lying excitations of the molecule are selected.
All determinants with at most a single hole in the Cu 3d shell and no doubly occupied Cu 4s orbital were included in \eqref{eq:sampling} to reflect the low-energy excitations observed in experiment.
Only the lowest energy state with a Cu 4s$^2$ electronic configuration was included, corresponding to a double excitation 2 O ${2p_\pi} \rightarrow$ 2 Cu 4s with an energy $>7$ eV above the UKS ground state.
Inclusion of this high energy determinant is necessary to ensure our model places similar Cu 4s$^2$ states high in energy.
No other Cu 4s$^2$ determinants were included since we were not interested in describing the variation among these high energy states.

A shell sampling scheme was used to sample states from the selected approximate low-energy space.
We began by fixing the coefficient $p_0 = \sqrt{w}$ where $w \in \{1.0, 0.8, ..., 0.2\}$. 
For each choice of $w$ we sampled n = 5 states by randomly selecting the unassigned coefficients from independent, standard normal distributions such that $\sum_j p_j^2 = 1$. 
This procedure generated a uniform set of samples from spherical shells at decreasing radii in determinant space centered at $|\text{D}_0\rangle$.
We repeated this process centered at every determinant in \eqref{eq:sampling} resulting in a sample set of overlapping shells which fill the space $\mathcal{LE}^\prime$.

The final sampled states were projected using FN-DMC to ensure they are as low in energy as possible.
The FN-DMC calculations were conducted with T-moves to ensure the method is variational while using pseudopotentials.
A timestep of $\tau = 0.01$ was used.

\subsubsection{Selecting candidate descriptors}
We select the following set of 1- and 2-body operators to construct our candidate descriptors based on our understanding of the low-energy excitations of the CuO molecule and our sampled states:
\begin{equation}
\begin{split}
\sum_{\sigma = \uparrow, \downarrow} \Bigg(\epsilon_{3d_\pi}\hat{n}_{3d_\pi,\sigma} + \epsilon_{3d_{z^2}}\hat{n}_{3d_{z^2},\sigma} +  \epsilon_{2p_z} \hat{n}_{2p_z,\sigma} + \epsilon_{2p_\pi}\hat{n}_{2p_\pi,\sigma} + \epsilon_{4s}\hat{n}_{4s,\sigma} +\\
t_\pi \hat{c}_{3d_\pi,\sigma}^\dagger \hat{c}_{2p_\pi,\sigma} + t_{dz} \hat{c}_{3d_{z^2},\sigma}^\dagger \hat{c}_{2p_z,\sigma}  + t_{sz}\hat{c}_{4s,\sigma}^\dagger \hat{c}_{2p_z,\sigma} + t_{sd}\hat{c}_{4s,\sigma}^\dagger \hat{c}_{3d_{z^2},\sigma} + \text{h.c.} + \Bigg)  \\
J_{sd}\sum_{i\in {\{\text{xy, xz, ...}}\}} \vec{S}_{4s} \cdot \vec{S}_{3d_i} + U_s \hat{n}_{4s,\uparrow}\hat{n}_{4s,\downarrow} + E_0.
\end{split}
\label{eq:models}
\end{equation}
The operator subscripts correspond to the intrinsic atomic orbital (IAO) basis elements constructed using the molecular orbitals in \eqref{eq:sampling}, and can be seen in \ref{fig:cuo_regr}a.
Operators are contracted over symmetry-related pairs for the $3d_\pi, 2p_\pi$ orbitals, for example $\hat{n}_{3d_\pi} = \hat{n}_{3d_{xz}} + \hat{n}_{3d_{yz}}$.

The parantheses of \eqref{eq:models} contain the full set of symmetry allowed 1-body operators within the Cu 3d, 4s and O 2p space.
These operators are required to describe the most basic single particle features seen in molecules like orbital occupations and hybridizations.
Two 2-body interaction terms are also considered: a Hund's coupling between Cu 4s and 3d orbitals and a Cu 4s Coulomb repulsion.
The prior term is motivated by the existence of a $\sim $ 0.5 eV Hund's coupling on the neutral copper atom.
Motivation for the latter comes from the FN-DMC energies of our sampled data.
We find that states with a Cu $3d^{10} 4s^{2}$ configuration lie at least 7 eV above our lowest energy sampled state, while the highest energy states with a $3d^{10} 4s^{1}$ configuration lie near 2 eV.

The operator $\hat{n}_{3d_\delta}$ is not included as it is linearly dependent on the rest of the occupation operators.
The linear dependency arises from the fixed number of electrons in the model.
As such, the operator $\hat{n}_{3d_\delta} = N_\text{elec} - \sum_{i \neq 3d_\delta} \hat{n}_i$.
The first term can be absorbed into the constant $E_0$, and the latter terms shift the occupation energies $\epsilon_i$.

\subsubsection{Fitting the model}
A direct attempt at fitting all eleven terms in \eqref{eq:models} leads to a high goodness of fit but unexpectedly large coefficients for the fit model.
The fit model had an R$^2= 0.99$, but the fit coefficients were very inconsistent with experimental measurements and our sampled data.
For example, the model predicted an enormous occupation energy gap of $\epsilon_{2p_\pi} - \epsilon_{3d_\delta} = $ 5.87 eV, while in experiment and our sampled data this scale was between 2 $-$ 3 eV.
This inflation of model parameters was found to be associated with multicollinearity among the candidate descriptors, quantified by a large condition number (CN) of 1470.

A principal component analysis (PCR) analysis was used to alleviate this multicollinearity and fit the effective theory.
PCR can be broken into three parts: conducting a principal component analysis (PCA) and selecting candidate principal components to work with (PCs), fitting a linear model to these PCs, and then inverting the coefficients back to the original descriptors.
If the subset of PCs are chosen appropriately, the final regressed model will have a similar goodness of fit to one fit on the original data matrix but without the consequence of serious multicollinearity.

From the PCA, the nine highest ranked PCs were selected to use in the PCR.
In Figure \eqref{fig:cuo_regr}b we plot both the cumulative explained variance ratio (CEVR) and CN of a matrix composed of the $N_\text{comp}$ highest ranked PCs varying from one to eleven.
The CEVR indicates how much of the variation in the descriptor values a subset of PCs capture, and the CN quantifies the multicollinearity among these PCs.
The objective in selecting PCs is to the maximize the prior while simultaneously minimizing the latter, ensuring a high goodness of fit with little to no multicollinearity.
Reducing the number of PCs from eleven rapidly decreases the CN to $\sim $ 200 while keeping the CEVR relatively fixed.
Further reducing the number of PCs past nine does not affect the CN but steadily decreases the CEVR, indicating that nine principal components best satisfies the objective above with a CEVR 0.9995 and CN of 257.

The model fit to these nine PCs maintains an excellent goodness of fit without serious multicollinearity.
When trained on all the sampled data this model yields an $R^2$ = 0.98 with a CN of 257. 
While the goodness of fit does decrease slightly compared to a fit with the full data matrix, the nearly six-fold reduced condition number readily compensates for this loss.
An inverse PCA transformation was used to map the fit coefficients corresponding to the PC descriptors back into the original eleven coefficients of \eqref{eq:models}.
These coefficients can be seen in Figure \ref{fig:cuo_regr}c with single standard deviation bootstrap error estimates.

\begin{figure*}
\centering
\includegraphics[width=1.0\linewidth]{./figs/cuo_regr.pdf}
\caption{Figures related to model regression for the CuO molecule: a) Intrinsic atomic orbitals used as a single particle basis for the model Hamiltonian, b) Results of PCA showing the CEVR and CN of a data set containing N$_{comp}$  highest ranked PCs, c) The final regressed parameters from the PCR.}
\label{fig:cuo_regr}
\end{figure*}

\subsubsection{Solutions of the model}
The model is solved by exact diagonalization, and the solutions agree well with experimental spectra and state assignments.
Shown in Figure \ref{fig:comp} is a comparison between the excitation energies calculated from our model and those seen in the latest anion photoelectron spectroscopy measurements.
The model energies for these eigenstates are denoted by orange circles and agree with experiment up to a maximum deviation of 0.2 eV.
The term symbols assigned in experiment also match the term symbols of the model eigenstates, which are constructed by looking at the multiplicity and electronic configuration of each model eigenstate.

\begin{figure*}
\centering
\includegraphics{./figs/comp.pdf}
\caption{Comparison of the four lowest lying eigenstates of the CuO moelcule between experiment and three computational methods. The model and UKS calculations are from this work. Text near the data points corresponds to the molecular term symbols of the eigenstates.}
\label{fig:comp}
\end{figure*}

The model states also agree with recent theoretical calculations like UKS and CCSD(T).
The energies for the UKS and CCSD(T) states are shown in Figure \ref{fig:comp} as green and red circles, respectively.
The UKS energies were calculated in this work in en route to constructing the determinants $|\text{D}_j\rangle$ used in our sampling scheme \eqref{eq:sampling}.
Our model energies agree with both theoretical techniques within errorbars for the first three states shown.
The energy of the highest energy $^2\Delta$ is severely over estimated using both UKS and CCSD(T), by nearly 1 eV.
This over estimation may arise from an inadequate relaxation of the $\sigma$ type orbitals which form a three orbital hybridization complex in this system.

@@

The first system I will develop a model for is the benzene molecule.
The benzene molecule has chemical formula $C_6 H_6$ and contains a hexagonal ring of carbon atoms with hydrogen atoms sticking out from each carbon.
These carbon atoms are linked by sp$^2$ hybridized $\sigma$ bonds and an alternating sequence of $\pi$ bonds from the C 2p$_z$ orbitals.
Intuitively, the ground state of the system can be seen as a superposition of two configurations of alternating $\pi$ bonds as seen in Figure \ref{fig:proposed}a rotated 30$^o$ relative to each other.

\begin{figure*}
\centering
\includegraphics[width=1.0\linewidth]{./figs/proposed.pdf}
\caption{Schematic diagrams for the a) benzene molecule, b) AB and AA stacked BLG \textbf{ref}.}
%Voltage Tunable Plasmon Propagation in Dual Gated Bilayer Graphene
\label{fig:proposed}
\end{figure*}

@@

The last systems I will develop models for are AA and AB stacked BLG.
AA stacked BLG is simply two sheets of graphene placed on top of one another where each atom is aligned with the atom beneath it.
In AB stacked BLG, a carbon atom from the lower layer is present in the center of each carbon hexagon of the upper layer, and vice versa.
As such, one can think of AB BLG as a shifted version of AA BLG, where the shift is half the size of a single carbonic hexagon.
Schematic drawings of these two stackings can be seen in Figure \ref{fig:proposed}b.
